---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r dataload, echo=TRUE, message=FALSE, results="hide"}
require(dplyr)
require(lattice)
require(rmarkdown)
stepData <- read.csv(".\\activity.csv", header = TRUE, colClasses = c(NA, "Date", NA))
```

## What is mean total number of steps taken per day?
```{r totalsteps_nofill, echo=TRUE}
stepDataByDay <- summarize(group_by(stepData, date), steps = sum(steps, na.rm=TRUE))
histogram(~steps, data = stepDataByDay, breaks = 10, type = "count",
          xlab = "Total Number of Steps Taken per Day")
stepDataByDayMean <- round(mean(stepDataByDay$steps))
stepDataByDayMedian <- round(median(stepDataByDay$steps))
```

The mean total number of steps taken per day is `r toString(stepDataByDayMean)`.

The median total number of steps taken per day is `r toString(stepDataByDayMedian)`.

## What is the average daily activity pattern?

```{r averagesteps_nofill, echo=TRUE}
stepDataByInterval <- summarize(group_by(stepData, interval), steps = mean(steps, na.rm=TRUE))
xyplot(steps~interval, data = stepDataByInterval, type="l",
     xlab = "Interval",
     ylab = "Mean Number of Steps Taken")
stepDataByIntervalWhichMax <- stepDataByInterval$interval[which.max(stepDataByInterval$steps)]
```

The interval with the maximum number of steps taken per day is `r stepDataByIntervalWhichMax`.

## Imputing missing values

```{r missingvalues, echo=TRUE}
countMissingValues <- sum(!complete.cases(stepData))
```

The number of missing cases in the dataset is `r countMissingValues`.

Our strategy to fill in all the missing values will be to take the mean of values for that 5-minute interval.

```{r datafill, echo=TRUE}
stepDataFilled <- stepData %>%
    mutate(steps = if_else(is.na(steps), stepDataByInterval[interval/5+1,2][[1]], as.double(steps)))
```

```{r totalsteps_fill, echo=TRUE}
stepDataFilledByDay <- summarize(group_by(stepDataFilled, date), steps = sum(steps, na.rm=TRUE))
histogram(~steps, data = stepDataFilledByDay, breaks = 10, type = "count",
          xlab = "Total Number of Steps Taken per Day")
stepDataFilledByDayMean <- round(mean(stepDataFilledByDay$steps))
stepDataFilledByDayMedian <- round(median(stepDataFilledByDay$steps))
```

The mean total number of steps taken per day is `r toString(stepDataFilledByDayMean)`. This is higher than the same number without data filling, which is expected since NA values were counted as 0 without filling and as a positive value with filling.

The median total number of steps taken per day is `r toString(stepDataFilledByDayMedian)`. This is the exact same number as without data filling.

The distribution represented in the histogram with data filling looks more like a normal distribution, whereas the one without data filling had a strong skew due to 0 values.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays, echo=TRUE}
weekDays <- weekdays(stepDataFilled$date)
stepDataFilled$day <- as.factor(
    if_else(weekDays == "Saturday" | weekDays == "Sunday",
            "weekend", "weekday"))
```

```{r averagesteps_fill, echo=TRUE}
stepDataFilledByInterval <- summarize(group_by(stepDataFilled, interval, day), steps = mean(steps, na.rm=TRUE))
xyplot(steps~interval|day, data = stepDataFilledByInterval, type="l",
     xlab = "Interval",
     ylab = "Mean Number of Steps Taken",
     layout = c(1,2))
```